Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
more rearrangements

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on gravity theories that deviate
from general relativity through higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against the two loudest
gravitational-wave ringdown signals observed so far, notably, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the coupling
constants of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Since the first detection of gravitational waves (GWs) from a binary black-hole (BBH) merger in 2015~\cite{LIGOScientific:2016aoc},
the LIGO~\cite{LIGOScientific:2014pky} and Virgo~\cite{VIRGO:2014yos} detectors have observed about 90 GWs signals \cite{LIGOScientific:2021djp} from mergers
of BHs, neutron stars (NSs) \cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki,LIGOScientific:2020aai} and their mixture \cite{LIGOScientific:2021qlt}. These results have been confirmed
by independent analyses, which have also identified a few new GW signals~\cite{Nitz:2018imz,Nitz:2019hdf,Venumadhav:2019lyq,Zackay:2019btq,Nitz:2021zwj,Olsen:2022pin}.

@@

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the intermediate (plunge) and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow to set upper limit on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. Different line colors distinguish between events, while
different line styles distinguish between different $N_{\rm max}$.
%
We see that this parameter which controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR.
The right panel shows the PDFs in the range $0 \leqslant \gamma_{\rm EdGB} \leqslant 1$. Note how
the curves are flat in this range for both events and when $N_{\rm max} = 1$.}
\label{fig:gamma_gb}
\end{figure}

@@

@@

\begin{table}[b]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event &  EFT    & ParSpec & Constraint      \\
              &       &  bound? & bound?  & ($\gm = M_{f}$) \\
\hline
  & GW150914 & No  & Yes & -- \\
0 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
  & GW150914 & No  & Yes & -- \\
1 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for EdGB gravity for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We find that we cannot place any constraint on $\ell_{\rm EdGB}$ with our waveform model from either GW event.
}
\label{tab:summary_edgb}
\end{table}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on theories that deviate
from general relativity by higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against the two loudest
gravitational-wave ringdown signals observed so far, notably, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the coupling
constants of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

Since the first detection of gravitational waves (GWs) from a binary black-hole (BBH) merger in 2015~\cite{LIGOScientific:2016aoc},
the LIGO~\cite{LIGOScientific:2014pky} and Virgo~\cite{VIRGO:2014yos} detectors have observed about 90 GWs signals~\cite{LIGOScientific:2021djp} from mergers
of BHs, neutron stars (NSs) \cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki,LIGOScientific:2020aai} and their mixture~\cite{LIGOScientific:2021qlt}. These results have been confirmed
by independent analyses, which have also identified a few new GW signals~\cite{Nitz:2018imz,Nitz:2019hdf,Venumadhav:2019lyq,Zackay:2019btq,Nitz:2021zwj,Olsen:2022pin}.

@@

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow to set upper limit on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. Different line colors distinguish between events, while
different line styles distinguish between different $N_{\rm max}$.
%
We see that this parameter which controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR.
The right panel shows the PDFs in the range $0 \leqslant \gamma_{\rm EdGB} \leqslant 1$. Note how
the curves are flat in this range for both events and when $N_{\rm max} = 1$.}
\label{fig:gamma_gb}
\end{figure}

@@

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event &  EFT    & ParSpec & Constraint      \\
              &       &  bound? & bound?  & ($\gm = M_{f}$) \\
\hline
  & GW150914 & No  & Yes & -- \\
0 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
  & GW150914 & No  & Yes & -- \\
1 & GW200129 & No  & Yes & -- \\
  & Combined & --  & Yes & -- \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for EdGB gravity for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We find that we cannot place any constraint on $\ell_{\rm EdGB}$ with our waveform model from either GW event.
}
\label{tab:summary_edgb}
\end{table}
